---
title: "95% bootstrap confidence intervals for the differences in medians"
author: "Maksim Helmann"
format: html
editor: visual
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: load-packages
#| include: false
#| warning: false
source("../.Rprofile")
```

```{r}
#| echo: false
finaldata <- read.csv(here("data", "analytical", "final_mort_data.csv"), header = TRUE)
finaldata$loggdp <- log(finaldata$gdp1000)
```

```{r}
data2017.inf <- finaldata |>
  dplyr::filter(year == 2017) |>
  dplyr::filter(!is.na(infmor)) 

data2017.un5 <- finaldata |>
  dplyr::filter(year == 2017) |>
  dplyr::filter(!is.na(un5mor)) 

data2017.neo <- finaldata |>
  dplyr::filter(year == 2017) |>
  dplyr::filter(!is.na(neomor)) 

```

Compute bootstrap samples calculate confidence intervals from the bootstrap samples:

```{r}
library(boot)
# Infant mortality
set.seed(2024)
getmeddiff <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$infmor, sample_data$armconf1, FUN = function(x) median(x,na.rm=TRUE))
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
  }
bootout <- boot(data2017.inf, statistic = getmeddiff, strata = data2017.inf$armconf1, R = 1000)
inf.ci <- boot.ci(boot.out = bootout, conf = 0.95, type = c("basic", "perc", "bca"))


set.seed(2024)
getmeddiff <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$un5mor, sample_data$armconf1, FUN = function(x) median(x,na.rm=TRUE))
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
  }
bootout <- boot(data2017.un5, statistic = getmeddiff, strata = data2017.un5$armconf1, R = 1000)
un5.ci <- boot.ci(boot.out = bootout, conf = 0.95, type = c("basic", "perc", "bca"))


set.seed(2024)
getmeddiff <- function(data, indices) {
  sample_data <- data[indices, ]
  group_meds <- tapply(sample_data$neomor, sample_data$armconf1, FUN = function(x) median(x,na.rm=TRUE))
  meddiff <- group_meds[2] - group_meds[1]
  return(meddiff)
  }
bootout <- boot(data2017.neo, statistic = getmeddiff, strata = data2017.neo$armconf1, R = 1000)
neo.ci <- boot.ci(boot.out = bootout, conf = 0.95, type = c("basic", "perc", "bca"))

```

```{r}
inf.ci
```


```{r}
un5.ci
```


```{r}
neo.ci
```

In the following we will interpret the obtained 95% bootstrap confidence intervals for the difference in medians of infant mortality, under-5 mortality and neonatal mortality. The CI for all of them provide a significant evidence at $\alpha = 0.05$ that the difference in medians for countries with armed conflict and non armed conflict is greater than zero.




