---
title: "Maternal Mortality Trend"
author: "Maksim Helmann"
format: pdf
editor: visual
---


```{r}
#| echo: true
#| label: load-packages
#| include: false
source("../.Rprofile")
```

## Load data

```{r}
arm_conf_df <- read.csv(here("data", "analytical", "final_mort_data.csv"), 
                        header = TRUE)
```

The code below creates a new variable diffmatmor that shows the difference between maternal mortality of each year and maternal mortality in 2000. We then subselect the data where diffmatmor is greater than 0 for the year 2017 and save it in matmor_trend_df. This dataframe contains the data that will be used for the figure.
```{r}
count_w_conf <- arm_conf_df |>
  dplyr::select(country_name, ISO, year, matmor) |>
  dplyr::filter(year < 2018) |>
  arrange(ISO, year) |>
  group_by(ISO) |>
  mutate(diffmatmor = matmor - matmor[1L]) %>% 
  dplyr::filter(diffmatmor > 0 & year == 2017)

matmor_trend_df <- arm_conf_df %>% 
  dplyr::filter(ISO %in% count_w_conf$ISO)

```


## Mortality trend for 2000-2019 by OECD

The code below generates the mortality trend figure.
```{r}
#| warning: false
matmor_trend_fig <- matmor_trend_df |>
  ggplot(aes(x = year, y = matmor + 1, group = ISO, color = country_name)) +
  geom_line(alpha = 0.7) +
  xlim(c(2000,2017)) +
  scale_y_continuous(trans='log10') + 
  labs(y = "Maternal mortality (log 10 scale)", x = "Year", color = "Country", 
       title="Trend in maternal mortality for countries that
       had an increase from 2000 to 2017") + 
  theme_bw() +
  theme(plot.title = element_text(size=12, hjust = 0.5))

# Save the figure
ggsave(here("figures", "fig1_matmortrend.png"))
```


