---
title: "MI vs CC"
author: "Maksim Helmann"
format: html
editor: visual
execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| label: load-packages
#| include: false
#| warning: false
source("../.Rprofile")
```

```{r}
#| echo: false
finaldata <- read.csv(here("data", "analytical", "final_mort_data.csv"), header = TRUE)
finaldata$loggdp <- log(finaldata$gdp1000)
```

## CC

### Fit the four mortalita models using plm()
```{r}
#| echo: false
library(plm)
library(texreg)

finaldata.cc <- read.csv(here("data", "analytical", "final_mort_data.csv"), header = TRUE)
finaldata.cc$loggdp <- log(finaldata.cc$gdp1000)

preds <- as.formula(" ~ armconf1 + loggdp + OECD + popdens + urban + 
                  agedep + male_edu + temp + rainfall1000 + earthquake + drought + as.factor(ISOnum) + as.factor(year)")

matmormod <- plm(update.formula(preds, matmor ~ .), index = c("year", "ISO"), effect = "twoways", model = "within", data = finaldata.cc)
un5mormod <- plm(update.formula(preds, un5mor ~ .), index = c("year", "ISO"), effect = "twoways", model = "within", data = finaldata.cc)
infmormod <- plm(update.formula(preds, infmor ~ .), index = c("year", "ISO"), effect = "twoways", model = "within", data = finaldata.cc)
neomormod <- plm(update.formula(preds, neomor ~ .), index = c("year", "ISO"), effect = "twoways", model = "within", data = finaldata.cc)
```

## MI

Convert ISO to numeric
```{r}
midata <- finaldata |>
  mutate(ISOnum = as.numeric(as.factor(finaldata$ISO))) |>
  select(-country_name, -ISO)
```

Dry run to get meth and pred. Then edit meth and pred.
```{r}
#| echo: false
mice0  <- mice(midata, seed = 100, m = 5, maxit = 0, print = F)
meth <- mice0$method
meth[c("urban", "male_edu", "temp", "rainfall1000", "matmor", "infmor", "neomor", "un5mor", "loggdp", "popdens")] <- "2l.lmer"

pred <- mice0$predictorMatrix
pred[c("urban", "male_edu", "temp", "rainfall1000", "matmor", "infmor", "neomor", "un5mor", "loggdp", "popdens"), "ISOnum"] <- -2
```
Afterward, we perform MI with m=10 imputations.
```{r}
#| echo: false
#| output: false
mice.multi.out  <- mice(midata, seed = 100, m = 10, maxit = 10, # for rendering I only performed 10 iterations
                        method = meth,
                        predictorMatrix = pred)
```
Check for convergence
```{r}
#| echo: false
plot(mice.multi.out)
```
We can observe the all the plots show good convergence.
Now, we fit four regression models using the same 10 imputed data sets using the with() function.
```{r}
#| echo: false
mice.matmor <- with(data = mice.multi.out, 
                   expr = lm(matmor ~ armconf1 + loggdp + OECD + popdens + urban + 
                               agedep + male_edu + temp + rainfall1000 + earthquake + drought + as.factor(ISOnum) + as.factor(year)))

mice.neomor <- with(data = mice.multi.out, 
                   expr = lm(neomor ~ armconf1 + loggdp + OECD + popdens + urban + 
                               agedep + male_edu + temp + rainfall1000 + earthquake + drought + as.factor(ISOnum) + as.factor(year)))

mice.infmor <- with(data = mice.multi.out, 
                   expr = lm(infmor ~ armconf1 + loggdp + OECD + popdens + urban + 
                               agedep + male_edu + temp + rainfall1000 + earthquake + drought + as.factor(ISOnum) + as.factor(year)))

mice.un5mor <- with(data = mice.multi.out, 
                   expr = lm(un5mor ~ armconf1 + loggdp + OECD + popdens + urban + 
                               agedep + male_edu + temp + rainfall1000 + earthquake + drought +  as.factor(ISOnum) + as.factor(year)))

```

Use pool() function to summarize the simulation outcome.
```{r}
#| echo: false
mice.matmor.pool <- pool(mice.matmor)
mice.neomor.pool <- pool(mice.neomor)
mice.infor.pool <- pool(mice.infmor)
mice.un5mor.pool <- pool(mice.un5mor)
```


## Create a table that shows coefficients from CC analyses and MI analyses side by side

```{r}
#| echo: false

keepvars <- list("armconf1" = "Armed conflict",
                 "loggdp" = "log(GDP)",
                 "OECD" = "OECD",
                 "popdens" = "Population density",
                 "urban" = "Urban",
                 "agedep" = "Age dependency",
                 "male_edu" = "Male education",
                 "temp" = "Average temperature",
                 "rainfall" = "Average rainfall",
                 "earthquake" = "Earthquake",
                 "drought" = "Drought")

screenreg(list(matmormod, mice.matmor.pool,
               un5mormod, mice.un5mor.pool,
               infmormod, mice.infor.pool,
               neomormod, mice.neomor.pool),
          ci.force = TRUE,
          custom.coef.map = keepvars,
          custom.model.names = c("Mat Mort-CC", "Mat Mort-MI",
                                 "Neonatal Mort-CC", "Neonatal Mort-MI", 
                                 "Infant Mort-CC", "Infant Mort-MI",
                                 "Under-5 Mort-CC","Under-5 Mort-MI"))

```






